---
title: "PENPAL — Novelty, Transience, and Resonance (window-level, turn-agnostic)"
author: "PENPAL Collaboration Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 5)

library(tidyverse)
library(lme4)
library(lmerTest)
library(knitr)
library(broom)
library(gridExtra)

theme_set(theme_minimal(base_size = 13))

condition_colors <- c(
  "HH"      = "#2ca02c",  # Green
  "H-LLM"   = "#1f77b4",  # Blue
  "LLM-LLM" = "#ff7f0e"   # Orange
)
```

## 1. Overview & Setup

Surprisal-based **Novelty**, **Transience** and **Resonance** across HH, H-LLM and
LLM-LLM, computed on fixed word windows of the completed story with **no turn
structure**.

Two things differ from the EMNLP analysis and govern how everything below must
be read:

1. **The unit is a sub-turn window, not a turn.** At `w = 14` words (mean turn
   length in the source corpus is 26–28), a window sits mostly *inside* one
   author's contribution, and the "next window" is usually the same author
   continuing. Transience here therefore measures **local continuation**, not
   **uptake by a partner**. It is not the same construct as EMNLP's transience,
   and the innovation-bias slope is not comparable to EMNLP's turn-level number.
2. **The estimand must be within-story.** EMNLP modelled turns nested in stories
   (a within-story process). A story-level regression answers a different,
   between-story question and is reported here only as a diagnostic (§6).

```{r load-data}
root_dir <- if (basename(getwd()) == "analysis") ".." else "."

measures_path <- file.path(root_dir, "data", "processed", "surprisal", "penpal_measures_all.csv")
window_path   <- file.path(root_dir, "data", "processed", "surprisal", "surprisal_window_level.csv")
ann_path      <- file.path(root_dir, "data", "interim", "annotations", "penpal_annotations_final.csv")

condition_levels <- c("HH", "H-LLM", "LLM-LLM")

df_story <- read_csv(measures_path, show_col_types = FALSE) %>%
  mutate(cond = factor(cond, levels = condition_levels),
         llm  = as.numeric(cond) - 1)

df_ann <- if (file.exists(ann_path)) read_csv(ann_path, show_col_types = FALSE) else NULL

# The window file is REQUIRED for the primary model. Fail loudly rather than
# letting a downstream `if (!is.null(df_window))` silently skip the analysis --
# that is exactly how the mixed model went missing from earlier runs.
has_window <- file.exists(window_path)
if (!has_window) {
  warning(
    "MISSING: ", window_path, "\n",
    "The window-level mixed model (Section 5) is the primary analysis and cannot run.\n",
    "Regenerate it with:\n",
    "  python scripts/03_compute_surprisal_metrics.py --model google/gemma-4-31b --window 14\n",
    call. = FALSE
  )
}

df_window <- if (has_window) {
  read_csv(window_path, show_col_types = FALSE) %>%
    mutate(cond = factor(cond, levels = condition_levels))
} else NULL

cat("Story-level rows :", nrow(df_story), "\n")
cat("Window-level rows:", if (is.null(df_window)) "MISSING" else nrow(df_window), "\n")
if (!is.null(df_window)) {
  cat("Windows per story: mean",
      round(mean(table(df_window$id)), 1),
      "| range", paste(range(table(df_window$id)), collapse = "-"), "\n")
}
```

---

## 2. Human Evaluations & Inter-Annotator Reliability

```{r human-ratings}
dims <- c("consistency", "coherence", "originality", "surprisingness", "enjoyment", "quality")
out_cols <- c(dims, "overall")

# The measures table now covers the FULL corpus, including stories nobody has
# rated. Report both n's so the ratings are not read as covering all stories.
df_story %>%
  group_by(cond) %>%
  summarise(n_stories = n(),
            n_rated = sum(!is.na(overall)),
            across(all_of(intersect(out_cols, colnames(df_story))),
                   ~ round(mean(.x, na.rm = TRUE), 2))) %>%
  kable(caption = "Mean human ratings by condition (means over rated stories only)")
```

```{r mw-contrasts}
pairs <- list(c("H-LLM", "HH"), c("H-LLM", "LLM-LLM"), c("HH", "LLM-LLM"))

map_dfr(pairs, function(pair) {
  a <- pair[1]; b <- pair[2]
  p_vals <- map_dbl(dims, function(m) {
    va <- df_story %>% filter(cond == a) %>% pull(m) %>% na.omit()
    vb <- df_story %>% filter(cond == b) %>% pull(m) %>% na.omit()
    wilcox.test(va, vb)$p.value
  })
  tibble(Dimension = dims, Contrast = paste(a, "vs", b),
         p_value = p_vals, q_FDR = p.adjust(p_vals, method = "BH"))
}) %>%
  mutate(across(c(p_value, q_FDR), ~ round(.x, 4))) %>%
  kable(caption = "Mann-Whitney contrasts on human ratings (BH-FDR within contrast)")
```

### Reliability, pooled **and within condition**

Pooled reliability is inflated by between-condition variance: two annotators who
merely agree that HH stories are more original than LLM-LLM ones will show a
respectable pooled rho without agreeing about any individual story. The
within-condition columns are the honest ones. Absolute agreement is reported
alongside rho because correlation is depressed by range restriction, which bites
hard in LLM-LLM where several dimensions sit at ceiling.

```{r inter-annotator}
# The annotation file is WIDE: one row per story with ann1_* / ann2_* columns
# (there is no annotation_id / long format any more).
raw_dims <- c(
  consistency    = "coherence_element_consistency",
  coherence      = "coherence_logical_progression",
  originality    = "creativity_originality",
  surprisingness = "creativity_surprisingness",
  enjoyment      = "likeability_enjoyability",
  quality        = "likeability_quality"
)

if (!is.null(df_ann) && all(paste0("ann1_", raw_dims) %in% colnames(df_ann))) {

  ann_cond <- df_ann %>%
    mutate(cond = factor(recode(tolower(condition),
                                hh = "HH", ha = "H-LLM", aa = "LLM-LLM"),
                         levels = condition_levels))

  pick <- function(cond_filter = NULL) {
    map_dfr(names(raw_dims), function(d) {
      a <- suppressWarnings(as.numeric(ann_cond[[paste0("ann1_", raw_dims[[d]])]]))
      b <- suppressWarnings(as.numeric(ann_cond[[paste0("ann2_", raw_dims[[d]])]]))
      keep <- !is.na(a) & !is.na(b)
      if (!is.null(cond_filter)) keep <- keep & ann_cond$cond == cond_filter
      if (sum(keep) < 3) {
        return(tibble(Dimension = d, rho = NA_real_, mad = NA_real_, n = sum(keep)))
      }
      tibble(Dimension = d,
             rho = suppressWarnings(cor(a[keep], b[keep], method = "spearman")),
             mad = mean(abs(a[keep] - b[keep])),
             n   = sum(keep))
    })
  }

  rel_tbl <- pick() %>%
    select(Dimension, rho_pooled = rho, mad_pooled = mad, n_pooled = n) %>%
    left_join(pick("HH")      %>% select(Dimension, rho_HH = rho, mad_HH = mad), by = "Dimension") %>%
    left_join(pick("H-LLM")   %>% select(Dimension, rho_HLLM = rho, mad_HLLM = mad), by = "Dimension") %>%
    left_join(pick("LLM-LLM") %>% select(Dimension, rho_AA = rho, mad_AA = mad), by = "Dimension") %>%
    mutate(across(where(is.numeric), ~ round(.x, 3)))

  kable(rel_tbl,
        caption = paste("Inter-annotator reliability, pooled vs within condition.",
                        "rho = Spearman; mad = mean |ann1 - ann2| on the 5-point scale.",
                        "Pooled rho is inflated by between-condition variance;",
                        "mad is robust to the ceiling effects in LLM-LLM."))
} else {
  message("Annotation file does not have the expected ann1_/ann2_ columns; skipping reliability.")
}
```

---

## 3. NTR descriptives by condition

```{r ntr-descriptives}
surp_cols <- c("s_novelty", "s_transience", "s_resonance")

df_story %>%
  group_by(cond) %>%
  summarise(n = n(),
            across(all_of(surp_cols),
                   list(mean = ~ round(mean(.x, na.rm = TRUE), 3),
                        sd   = ~ round(sd(.x, na.rm = TRUE), 3)),
                   .names = "{.col}_{.fn}")) %>%
  kable(caption = "Surprisal NTR by condition (bits). Less negative = more novel / more transient.")
```

## 4. Trend with model involvement (RQ1)

The simple ordinal trend across HH < H-LLM < LLM-LLM. This is the most robust
result in the surprisal family and is reported independently of any slope model.

```{r llmness-trend}
map_dfr(surp_cols, function(m) {
  ct <- suppressWarnings(cor.test(df_story$llm, df_story[[m]],
                                  method = "spearman", exact = FALSE))
  kw <- kruskal.test(df_story[[m]] ~ df_story$cond)
  tibble(Metric = m,
         rho_LLMness = round(unname(ct$estimate), 3),
         p_trend = signif(ct$p.value, 3),
         KW_chisq = round(unname(kw$statistic), 2),
         p_KW = signif(kw$p.value, 3))
}) %>%
  mutate(q_FDR = signif(p.adjust(p_trend, method = "BH"), 3)) %>%
  kable(caption = "Monotone trend with model involvement (Spearman on ordinal condition) + Kruskal-Wallis")
```

```{r ntr-contrasts}
map_dfr(pairs, function(pair) {
  a <- pair[1]; b <- pair[2]
  p_vals <- map_dbl(surp_cols, function(m) {
    va <- df_story %>% filter(cond == a) %>% pull(m) %>% na.omit()
    vb <- df_story %>% filter(cond == b) %>% pull(m) %>% na.omit()
    wilcox.test(va, vb)$p.value
  })
  tibble(Metric = surp_cols, Contrast = paste(a, "vs", b),
         p_value = signif(p_vals, 3),
         q_FDR = signif(p.adjust(p_vals, method = "BH"), 3))
}) %>%
  kable(caption = "Pairwise condition contrasts on NTR (BH-FDR within contrast)")
```

---

## 5. Innovation bias — window-level mixed models (primary)

### 5.1 Why the predictor must be decomposed

`s_resonance` is by construction `s_novelty - s_transience`. Regressing
resonance on novelty therefore has the predictor inside the outcome, and the
slope is an exact reparameterisation:

$$\beta_{\text{res}\sim\text{nov}} \;=\; 1 - \beta_{\text{tra}\sim\text{nov}}$$

So "innovation bias > 1" is *identically* the statement "more novel windows have
**lower** (more negative) transience", i.e. stronger forward binding. Both
parameterisations are reported; the transience model is the transparent one.

A raw `s_novelty` predictor also confounds two different comparisons: differences
*between* stories (some stories are globally less predictable) and variation
*within* a story (this window is more surprising than that story's norm). EMNLP
estimated the within-story process. We therefore split novelty into a
story-mean (between) and a story-centred deviation (within).

```{r window-prep, eval = has_window}
df_win <- df_window %>%
  filter(!is.na(s_resonance), !is.na(s_novelty), !is.na(s_transience), !is.na(cond)) %>%
  group_by(id) %>%
  mutate(nov_story_mean = mean(s_novelty, na.rm = TRUE),
         nov_within     = s_novelty - nov_story_mean) %>%
  ungroup() %>%
  mutate(nov_story_mean_c = as.numeric(scale(nov_story_mean, scale = FALSE)))

cat("Windows:", nrow(df_win), "in", n_distinct(df_win$id), "stories\n")
cat("Within-story SD of novelty:", round(sd(df_win$nov_within), 3),
    "| between-story SD:", round(sd(df_win$nov_story_mean), 3), "\n")
```

### 5.2 Primary model — transience on within-story novelty

```{r window-model-transience, eval = has_window}
m_tra <- lmer(
  s_transience ~ nov_within * cond + nov_story_mean_c + (1 + nov_within | id),
  data = df_win,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

# Fall back to a random intercept if the random-slope model is singular.
if (isSingular(m_tra, tol = 1e-4)) {
  message("Random-slope model singular; refitting with random intercept only.")
  m_tra <- lmer(
    s_transience ~ nov_within * cond + nov_story_mean_c + (1 | id),
    data = df_win,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
  )
}

summary(m_tra)
anova(m_tra)
```

### 5.3 Same model in the resonance parameterisation

Identical fit, rescaled; included so the slopes are directly comparable to how
EMNLP reported them.

```{r window-model-resonance, eval = has_window}
m_res <- update(m_tra, s_resonance ~ .)
summary(m_res)
```

### 5.4 Per-condition slopes

```{r window-slopes, eval = has_window}
if (requireNamespace("emmeans", quietly = TRUE)) {
  library(emmeans)
  emm_options(lmerTest.limit = 10000, pbkrtest.limit = 10000)
  tr_slopes  <- emtrends(m_tra, ~ cond, var = "nov_within")
  res_slopes <- emtrends(m_res, ~ cond, var = "nov_within")

  as.data.frame(tr_slopes) %>%
    rename(any_of(c(lower.CL = "asymp.LCL", upper.CL = "asymp.UCL"))) %>%
    select(cond, transience_slope = nov_within.trend, SE, lower.CL, upper.CL) %>%
    left_join(as.data.frame(res_slopes) %>%
                rename(any_of(c(lower.CL = "asymp.LCL", upper.CL = "asymp.UCL"))) %>%
                select(cond, resonance_slope = nov_within.trend), by = "cond") %>%
    mutate(check_1_minus = round(1 - transience_slope, 3),
           across(where(is.numeric), ~ round(.x, 3))) %>%
    kable(caption = "Within-story slopes by condition. resonance_slope should equal 1 - transience_slope.")

  cat("\nPairwise contrasts of the within-story novelty slope:\n")
  print(pairs(tr_slopes, adjust = "BH"))
} else {
  message("Install `emmeans` for per-condition slopes.")
}
```

---

## 6. Diagnostic: why the story-level model is NOT reported

The previous version of this document fitted `lm(s_resonance ~ s_novelty * cond)`
on 147 story means. That model is retained here **only** to show why it cannot
be interpreted as innovation bias.

At story level the slope is again `1 - beta(transience ~ novelty)`, and between
stories novelty and transience are strongly *positively* coupled: both are built
from the same contextual-surprisal term, and both mostly track a single nuisance
dimension — how predictable the story is to the LM overall. Because that
coupling is positive, every story-level slope is forced below 1. And because the
coupling is *strongest* in the most stylistically homogeneous condition, the
slope ordering simply inverts within-condition homogeneity. It carries no
information about narrative uptake.

```{r story-level-diagnostic}
diag_tbl <- df_story %>%
  filter(!is.na(s_novelty), !is.na(s_transience), !is.na(s_resonance)) %>%
  group_by(cond) %>%
  summarise(
    n              = n(),
    slope_res_nov  = round(coef(lm(s_resonance  ~ s_novelty))[2], 3),
    slope_tra_nov  = round(coef(lm(s_transience ~ s_novelty))[2], 3),
    one_minus_tra  = round(1 - coef(lm(s_transience ~ s_novelty))[2], 3),
    r_nov_tra      = round(cor(s_novelty, s_transience), 3),
    sd_novelty     = round(sd(s_novelty), 3),
    .groups = "drop"
  )
kable(diag_tbl,
      caption = "Story-level (BETWEEN-story) slopes. slope_res_nov tracks 1 - slope_tra_nov; the ordering follows r(nov,tra), i.e. within-condition homogeneity, not uptake.")

gap <- with(df_story, abs(s_resonance - (s_novelty - s_transience)))
cat("Identity check  max |s_resonance - (s_novelty - s_transience)| =",
    round(max(gap, na.rm = TRUE), 4), "\n")
cat("If this exceeds ~1e-6, the three story means were averaged over different\n",
    "window index ranges -- regenerate with the fixed scripts/03_compute_surprisal_metrics.py\n")
```

---

## 7. Correlations with human judgments

Reported for the three NTR measures and the stylometric block, with FDR across
all tested pairs. Treat as exploratory: the confirmatory claims are the
condition trends in §4.

```{r correlations}
style_cols <- c("n_words", "n_sentences", "mean_word_length", "mean_sentence_length",
                "ttr", "hapax_ratio", "comma_density", "flesch_approx",
                "adjacent_sentence_similarity", "semantic_distinctiveness")
existing_meas <- intersect(c(style_cols, surp_cols), colnames(df_story))

corr_long <- expand_grid(Metric = existing_meas, Rating = dims) %>%
  mutate(res = map2(Metric, Rating, function(m, d) {
    v <- df_story[[m]]; r <- df_story[[d]]
    ok <- !is.na(v) & !is.na(r)
    if (sum(ok) < 6) return(tibble(rho = NA_real_, p = NA_real_))
    ct <- suppressWarnings(cor.test(v[ok], r[ok], method = "spearman", exact = FALSE))
    tibble(rho = unname(ct$estimate), p = ct$p.value)
  })) %>%
  unnest(res) %>%
  mutate(q_FDR = p.adjust(p, method = "BH"))

corr_long %>%
  select(Metric, Rating, rho) %>%
  pivot_wider(names_from = Rating, values_from = rho) %>%
  mutate(across(-Metric, ~ round(.x, 2))) %>%
  kable(caption = "Spearman rho: computational metrics x human ratings")

corr_long %>%
  filter(!is.na(q_FDR), q_FDR < .05) %>%
  arrange(q_FDR) %>%
  mutate(rho = round(rho, 3), q_FDR = signif(q_FDR, 3)) %>%
  select(Metric, Rating, rho, q_FDR) %>%
  kable(caption = "Surviving FDR < .05")
```

---

## 8. Visualizations

```{r fig-eval-profile, fig.cap="Human evaluation profile by condition"}
df_story %>%
  pivot_longer(all_of(dims), names_to = "Dimension", values_to = "Rating") %>%
  mutate(Dimension = factor(Dimension, levels = dims)) %>%
  group_by(cond, Dimension) %>%
  summarise(Mean_Rating = mean(Rating, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(Dimension, Mean_Rating, color = cond, group = cond)) +
  geom_line(linewidth = 1.2) + geom_point(size = 3) +
  scale_color_manual(values = condition_colors) +
  labs(y = "Mean rating (1-5)", color = "Condition",
       title = "Human evaluation profile by condition") +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

```{r fig-surprisal-ntr, fig.cap="Surprisal NTR by condition"}
df_story %>%
  pivot_longer(all_of(surp_cols), names_to = "Metric", values_to = "Value") %>%
  mutate(Metric = factor(Metric, levels = surp_cols,
                         labels = c("Novelty", "Transience", "Resonance"))) %>%
  group_by(cond, Metric) %>%
  summarise(Mean = mean(Value, na.rm = TRUE),
            SE = sd(Value, na.rm = TRUE) / sqrt(n()), .groups = "drop") %>%
  ggplot(aes(cond, Mean, color = cond)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = Mean - 1.96 * SE, ymax = Mean + 1.96 * SE),
                width = 0.2, linewidth = 0.9) +
  facet_wrap(~ Metric, scales = "free_y") +
  scale_color_manual(values = condition_colors) +
  labs(x = "Condition", y = "Surprisal (bits)", color = "Condition",
       title = "Surprisal NTR across conditions")
```

### Within-story innovation bias

Each thin line is one story's own within-story relationship; thick lines are the
condition-level fits. This is the estimand the mixed model targets — unlike a
scatter of story means, which shows only between-story variation.

```{r fig-within-slopes, eval = has_window, fig.cap="Within-story transience vs novelty", fig.height=4.2}
ggplot(df_win, aes(nov_within, s_transience, color = cond)) +
  geom_point(alpha = 0.12, size = 0.7) +
  geom_smooth(aes(group = id), method = "lm", se = FALSE,
              linewidth = 0.25, alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.4, color = "black") +
  facet_wrap(~ cond) +
  scale_color_manual(values = condition_colors, guide = "none") +
  labs(x = "Novelty (within-story centred, bits)", y = "Transience (bits)",
       title = "Within-story slope of transience on novelty")
```

```{r fig-story-level-diagnostic, fig.cap="BETWEEN-story association (diagnostic only -- not innovation bias)"}
ggplot(df_story, aes(s_novelty, s_resonance, color = cond, fill = cond)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.2) +
  scale_color_manual(values = condition_colors) +
  scale_fill_manual(values = condition_colors) +
  labs(x = "Story-mean novelty (bits)", y = "Story-mean resonance (bits)",
       color = "Condition", fill = "Condition",
       title = "Between-story resonance vs novelty",
       subtitle = "Diagnostic only: slope = 1 - beta(transience~novelty), driven by nov/tra coupling")
```
