---
title: "PENPAL — Novelty, Transience, and Resonance (window-level, turn-agnostic)"
author: "PENPAL Collaboration Team"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    code-fold: show
    theme: flatly
    embed-resources: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 5)

library(tidyverse)
library(lme4)
library(lmerTest)
library(knitr)
library(broom)
library(gridExtra)
library(here)

theme_set(theme_minimal(base_size = 13))

condition_colors <- c(
  "HH"      = "#2ca02c",  # Green
  "H-LLM"   = "#1f77b4",  # Blue
  "LLM-LLM" = "#ff7f0e",   # Orange
  "LLM-LLM-cross" = "#d62728"
)
```

## 1. Theoretical Foundations & Corpus Setup

### 1.1 The Information-Theoretic Framework (Barron et al., 2018)

This analysis quantifies collaborative storytelling dynamics using information-theoretic measures of **Novelty ($N$)**, **Transience ($T$)**, and **Resonance ($R$)**, grounded in the foundational framework established by Barron, Huang, Spang, & DeDeo (2018; *PNAS*).

In an autoregressive causal language model, each token $w_t$ has an associated conditional probability $P(w_t \mid w_{<t})$ given the preceding context $w_{<t} = (w_1, \dots, w_{t-1})$. The **surprisal** (in bits, base-2 logarithm) is defined as:

$$S(w_t) = -\log_2 P(w_t \mid w_{<t})$$

Surprisal reflects unexpectedness: low surprisal denotes predictable, formulaic language conforming to the prior context, while high surprisal indicates novel vocabulary, sudden plot twists, or dramatic stylistic shifts.

Adapting the Barron et al. (2018) framework to sequential text generation:
1. **Novelty ($N$)**: Measures the informational departure of the current text window from the preceding narrative history. High novelty means the window introduces unexpected narrative elements not predicted by the story's past.
2. **Transience ($T$)**: Measures how quickly the information introduced in the current window is abandoned or forgotten in subsequent text. High transience indicates that the innovation was fleeting and failed to take root.
3. **Resonance ($R = N - T$)**: Quantifies the lasting, durable impact of the current window on the ongoing narrative trajectory. A positive resonance ($R > 0$) signifies that the novel ideas were successfully retained, adopted, and integrated into subsequent discourse; a negative resonance ($R < 0$) indicates that the window was an ephemeral distraction that quickly faded away.

### 1.2 Unit of Analysis: Canonical Sentence Units & Length Control

To capture fine-grained narrative dynamics across the entire story artifact, the active pipeline evaluates **canonical sentence units** (`surprisal_unit_level.csv`, $N = 9,202$ units across $N = 320$ stories across all four conditions):
- **Cross-Modal 1:1 Alignment**: Sentence segmentation is performed once via spaCy (`en_core_web_md`) and shared across the embedding trajectory, RQA, and surprisal pipelines (`sentence_index.csv`). Every sentence unit carries a universal `sent_idx` join key, allowing token-level predictive surprisal to be directly mapped to vector-space geometric excursions (`novelty_centroid`) and path detours (`path_detour`).
- **Controlling for Sentence Length**: Because mean sentence length varies systematically by condition (Human–Human: $\sim 13.8$ words; Human–AI: $\sim 17.4$ words; AI–AI: $\sim 22.4$ words), all sentence-level mixed-effects models explicitly carry sentence word length (`n_unit_words`) as a fixed covariate. This eliminates the length confounds that previously compromised topic-KL models while preserving natural syntactic and semantic boundaries.
- **Contrast with Fixed Sub-Turn Windows ($w = 14$)**: Earlier iterations evaluated fixed 14-word sliding windows to control word count by construction. However, arbitrary 14-word windows fragment sentences and dialogue quotation marks, and cannot be aligned to sentence-level recurrence matrices. Canonical sentences with statistical covariate adjustment provide the optimal synthesis of linguistic validity and cross-modal alignment.
- **Within-Story vs. Between-Story Estimands**: Crucially, narrative innovation is inherently a **within-story process** (whether a specific sentence is more surprising than this story's established norm). Story-level regressions conflate prompt difficulty with narrative innovation and are treated here strictly as a diagnostic (§6).

```{r load-data}

condition_levels <- c("HH", "H-LLM", "LLM-LLM", "LLM-LLM-cross")

measures_path <- here("data", "processed", "surprisal", "penpal_measures_all.csv")
window_path   <- here("data", "processed", "surprisal", "surprisal_unit_level.csv") # or window
ann_path      <- here("data", "interim", "annotations", "penpal_annotations_final.csv")


df_story <- read_csv(measures_path, show_col_types = FALSE) %>%
  mutate(cond = factor(cond, levels = condition_levels),
         llm  = if ("llm" %in% names(.)) llm else case_match(cond, "HH" ~ 0, "H-LLM" ~ 1, "LLM-LLM" ~ 2, "LLM-LLM-cross" ~ 2, .default = as.numeric(cond) - 1))

df_ann <- if (file.exists(ann_path)) read_csv(ann_path, show_col_types = FALSE) else NULL

# The window file is REQUIRED for the primary model. Fail loudly rather than
# letting a downstream `if (!is.null(df_window))` silently skip the analysis --
# that is exactly how the mixed model went missing from earlier runs.
has_window <- file.exists(window_path)
if (!has_window) {
  warning(
    "MISSING: ", window_path, "\n",
    "The window-level mixed model (Section 5) is the primary analysis and cannot run.\n",
    "Regenerate it with:\n",
    "  python scripts/03_compute_surprisal_metrics.py --model google/gemma-4-31b --window 14\n",
    call. = FALSE
  )
}

df_window <- if (has_window) {
  read_csv(window_path, show_col_types = FALSE) %>%
    mutate(cond = factor(cond, levels = condition_levels))
} else NULL

cat("Story-level rows :", nrow(df_story), "\n")
cat("Window-level rows:", if (is.null(df_window)) "MISSING" else nrow(df_window), "\n")
if (!is.null(df_window)) {
  story_win_counts <- df_window %>% count(id) %>% pull(n)
  cat("Windows per story: mean",
      round(mean(story_win_counts), 1),
      "| range", paste(range(story_win_counts), collapse = "-"), "\n")
}
```

---

## 2. Human Evaluations & Inter-Annotator Reliability

Independent human judges evaluated completed stories on six 5-point Likert dimensions:
1. **Consistency (`consistency`)**: Continuity of characters, entities, rules, and physical constraints across turns.
2. **Logical Progression (`coherence`)**: Smooth, sensible causality and thematic development over narrative time.
3. **Originality (`originality`)**: Creative inventiveness, freshness of premise, and avoidance of clichés.
4. **Surprisingness (`surprisingness`)**: Unpredictability of plot turns and unexpected twists.
5. **Enjoyment (`enjoyment`)**: Subjective reader pleasure and narrative immersion.
6. **Overall Quality (`quality`)**: Global aesthetic execution and literary merit.
- **Overall Score (`overall`)**: The composite unweighted mean across all evaluated dimensions.

> [!NOTE]
> **Evaluation Coverage**: Human ratings were gathered exclusively for the original three conditions (`HH`, $n = 36$; `H-LLM`, $n = 91$; `LLM-LLM`, $n = 20$; total $N = 147$). The simulated cross-model condition (`LLM-LLM-cross`, $n = 108$) was generated entirely by pairing distinct language models without human post-evaluation. Ratings analyses below are strictly evaluated on the annotated complete cases.

```{r human-ratings}
dims <- c("consistency", "coherence", "originality", "surprisingness", "enjoyment", "quality")
out_cols <- c(dims, "overall")

# The measures table covers the FULL corpus, including stories without human ratings.
# Report both n's so ratings are understood to cover only the rated subset.
df_story %>%
  group_by(cond) %>%
  summarise(n_stories = n(),
            n_rated = sum(!is.na(overall)),
            across(all_of(intersect(out_cols, colnames(df_story))),
                   ~ round(mean(.x, na.rm = TRUE), 2))) %>%
  kable(caption = "Mean human ratings by condition (means over rated stories only)")
```

### Interpretation of Mean Human Ratings
- **Coherence vs. Creativity Trade-off**: As shown in the table above, `LLM-LLM` stories achieved the highest ratings for logical consistency ($4.56$) and coherence ($4.45$), but scored lowest for originality ($2.75$) and surprisingness ($2.40$).
- **Human-Human Profile**: Human-Human stories (`HH`) show the inverse profile: lower consistency ($3.49$) and coherence ($3.26$), but higher originality ($3.44$) and surprisingness ($3.28$).
- **Human-AI Synergy**: Human-AI collaboration (`H-LLM`) strikes an optimal balance: high consistency ($4.22$) and coherence ($4.08$), coupled with solid originality ($3.23$) and the highest overall quality ($3.61$) and enjoyment ($3.65$).

```{r mw-contrasts}
rated_pairs <- list(c("H-LLM", "HH"), c("H-LLM", "LLM-LLM"), c("HH", "LLM-LLM"))

map_dfr(rated_pairs, function(pair) {
  a <- pair[1]; b <- pair[2]
  p_vals <- map_dbl(dims, function(m) {
    va <- df_story %>% filter(cond == a) %>% pull(m) %>% na.omit()
    vb <- df_story %>% filter(cond == b) %>% pull(m) %>% na.omit()
    wilcox.test(va, vb)$p.value
  })
  tibble(Dimension = dims, Contrast = paste(a, "vs", b),
         p_value = p_vals, q_FDR = p.adjust(p_vals, method = "BH"))
}) %>%
  mutate(across(c(p_value, q_FDR), ~ round(.x, 4))) %>%
  kable(caption = "Mann-Whitney contrasts on human ratings across rated conditions (BH-FDR within contrast)")
```

### Statistical Significance of Human Rating Contrasts
- **`H-LLM vs HH`**: Human-AI stories are significantly more coherent ($q_{\text{FDR}} = 0.0003$) and consistent ($q_{\text{FDR}} = 0.0004$) than pure human stories, with significantly higher overall quality ($q_{\text{FDR}} = 0.0069$).
- **`H-LLM vs LLM-LLM`**: Human-AI stories are significantly more original ($q_{\text{FDR}} = 0.0469$) and surprising ($q_{\text{FDR}} = 0.0152$) than pure AI stories, proving that human input is essential to disrupt formulaic model repetition.
- **`HH vs LLM-LLM`**: Human-Human stories show vastly superior surprisingness ($q_{\text{FDR}} = 0.0029$) and originality ($q_{\text{FDR}} = 0.0210$), but lower coherence ($q_{\text{FDR}} = 0.0002$) than same-model AI stories.

### Reliability: Pooled vs. Within-Condition Analysis

Reliability is assessed through two complementary statistics:
1. **Spearman Rank Correlation ($\rho$)**: Measures ordinal rank agreement between independent raters across stories.
2. **Mean Absolute Difference ($\text{MAD} = \frac{1}{n} \sum |\text{ann}_1 - \text{ann}_2|$)**: Quantifies absolute scale agreement in Likert score units on the 1–5 point scale.

> [!WARNING]
> **Methodological Nuance on Reliability**: Pooled $\rho$ is artificially inflated by between-condition variance (e.g., both annotators agree that HH stories are generally more original than LLM-LLM stories, producing high overall $\rho$). Conversely, within-condition $\rho$ is severely depressed by **range restriction**, particularly in `LLM-LLM` where ratings cluster tightly at the scale ceiling (4–5 for coherence). The MAD metric is robust to this ceiling effect and demonstrates strong absolute calibrational consensus ($\text{MAD} \approx 0.4\text{--}0.6$ points).
alongside rho because correlation is depressed by range restriction, which bites
hard in LLM-LLM where several dimensions sit at ceiling.

```{r inter-annotator}
# The annotation file is WIDE: one row per story with ann1_* / ann2_* columns
# (there is no annotation_id / long format any more).
raw_dims <- c(
  consistency    = "coherence_element_consistency",
  coherence      = "coherence_logical_progression",
  originality    = "creativity_originality",
  surprisingness = "creativity_surprisingness",
  enjoyment      = "likeability_enjoyability",
  quality        = "likeability_quality"
)

if (!is.null(df_ann) && all(paste0("ann1_", raw_dims) %in% colnames(df_ann))) {

  ann_cond <- df_ann %>%
    mutate(cond = factor(recode(tolower(condition),
                                hh = "HH", ha = "H-LLM", aa = "LLM-LLM", aa_cross="LLM-LLM-cross"),
                         levels = condition_levels))

  pick <- function(cond_filter = NULL) {
    map_dfr(names(raw_dims), function(d) {
      a <- suppressWarnings(as.numeric(ann_cond[[paste0("ann1_", raw_dims[[d]])]]))
      b <- suppressWarnings(as.numeric(ann_cond[[paste0("ann2_", raw_dims[[d]])]]))
      keep <- !is.na(a) & !is.na(b)
      if (!is.null(cond_filter)) keep <- keep & ann_cond$cond == cond_filter
      if (sum(keep) < 3) {
        return(tibble(Dimension = d, rho = NA_real_, mad = NA_real_, n = sum(keep)))
      }
      tibble(Dimension = d,
             rho = suppressWarnings(cor(a[keep], b[keep], method = "spearman")),
             mad = mean(abs(a[keep] - b[keep])),
             n   = sum(keep))
    })
  }

  rel_tbl <- pick() %>%
    select(Dimension, rho_pooled = rho, mad_pooled = mad, n_pooled = n) %>%
    left_join(pick("HH")      %>% select(Dimension, rho_HH = rho, mad_HH = mad), by = "Dimension") %>%
    left_join(pick("H-LLM")   %>% select(Dimension, rho_HLLM = rho, mad_HLLM = mad), by = "Dimension") %>%
    left_join(pick("LLM-LLM") %>% select(Dimension, rho_AA = rho, mad_AA = mad), by = "Dimension") %>%
    mutate(across(where(is.numeric), ~ round(.x, 3)))

  kable(rel_tbl,
        caption = paste("Inter-annotator reliability, pooled vs within condition.",
                        "rho = Spearman; mad = mean |ann1 - ann2| on the 5-point scale.",
                        "Pooled rho is inflated by between-condition variance;",
                        "mad is robust to the ceiling effects in LLM-LLM."))
} else {
  message("Annotation file does not have the expected ann1_/ann2_ columns; skipping reliability.")
}
```

### Detailed Interpretation of Annotator Reliability
- **Pooled Reliability ($\rho_{\text{pooled}} \approx 0.53\text{--}0.68$)**: Across the pooled corpus, annotator agreement is highest for `consistency` ($\rho = 0.68$) and `coherence` ($\rho = 0.65$), reflecting clear alignment on narrative logical structure.
- **Within-Condition Variance & Range Restriction**: When isolating conditions, $\rho$ drops markedly in `LLM-LLM` (e.g. $\rho = 0.28$ for coherence). This drop is **not** due to annotator disagreement, but rather due to severe **range restriction**: both annotators assign nearly all `LLM-LLM` stories scores of 4 or 5, compressing the score variance and depressing correlation coefficients.
- **Scale Agreement via MAD**: The mean absolute difference ($\text{MAD}$) remains steady at $0.38\text{--}0.56$ points across all dimensions and conditions. On a 5-point Likert scale, an average discrepancy of approximately half a point confirms excellent calibrational consensus.

---

## 3. NTR Descriptives by Condition

The table below reports story-level means and standard deviations for Surprisal Novelty ($S_N$), Transience ($S_T$), and Resonance ($S_R = S_N - S_T$) in bits.

> [!NOTE]
> **Surprisal Polarity & Units**: By mathematical construction of surprisal differencing, values are negative. **Less negative values correspond to higher novelty and higher transience**. Resonance represents the net difference ($S_N - S_T$).

```{r ntr-descriptives}
surp_cols <- c("s_novelty", "s_transience", "s_resonance")

df_story %>%
  group_by(cond) %>%
  summarise(n = n(),
            across(all_of(surp_cols),
                   list(mean = ~ round(mean(.x, na.rm = TRUE), 3),
                        sd   = ~ round(sd(.x, na.rm = TRUE), 3)),
                   .names = "{.col}_{.fn}")) %>%
  kable(caption = "Table: Surprisal NTR by condition (bits). Less negative = more novel / more transient.")
```

### Detailed Interpretation of NTR Descriptives
- **Novelty ($S_N$)**: Human-Human stories (`HH`: mean $-1.48$, $SD = 0.28$) exhibit the highest variance in novelty, reflecting unconstrained human authorial creativity. Human-AI (`H-LLM`: $-1.49$) matches this level of novelty while maintaining lower variance ($SD = 0.20$). AI-AI stories (`LLM-LLM`: $-1.46$, `LLM-LLM-cross`: $-1.47$) exhibit slightly higher mean surprisal values with compact standard deviations.
- **Transience ($S_T$)**: Transience reflects how quickly information introduced in a window fades. Pure AI pairs (`LLM-LLM`: $-1.43$) show slightly lower transience than cross-model AI pairs (`LLM-LLM-cross`: $-1.40$), indicating that same-model self-continuations sustain their own narrative threads longer.
- **Resonance ($S_R = S_N - S_T$)**: Because story-level means aggregate across the entire text, net resonance hovers near zero across all conditions (means between $-0.03$ and $-0.07$ bits). The true operational divergence between conditions emerges at the **within-story window level**, where local innovation bias operates (§5).

---

## 4. Trend with Model Involvement & Condition Contrasts (RQ1)

Model involvement tracks the number of artificial intelligence participants in the collaborative dyad:
- `0`: Human-Human (`HH`)
- `1`: Human-AI (`H-LLM`)
- `2`: Same-Model AI-AI (`LLM-LLM`) and Cross-Model AI-AI (`LLM-LLM-cross`)

We evaluate two distinct statistical questions:
1. **Monotone Trend**: Does increasing AI involvement systematically alter surprisal dynamics (tested via Spearman rank correlation $\rho_{\text{LLMness}}$)?
2. **Omnibus & Pairwise Differences**: Do the four conditions differ non-parametrically across the board (Kruskal-Wallis $\chi^2$), and specifically does cross-model AI collaboration (`LLM-LLM-cross`) differ from same-model AI collaboration (`LLM-LLM`)?

```{r llmness-trend}
map_dfr(surp_cols, function(m) {
  ct <- suppressWarnings(cor.test(df_story$llm, df_story[[m]],
                                  method = "spearman", exact = FALSE))
  kw <- kruskal.test(df_story[[m]] ~ df_story$cond)
  tibble(Metric = m,
         rho_LLMness = round(unname(ct$estimate), 3),
         p_trend = signif(ct$p.value, 3),
         KW_chisq = round(unname(kw$statistic), 2),
         p_KW = signif(kw$p.value, 3))
}) %>%
  mutate(q_FDR = signif(p.adjust(p_trend, method = "BH"), 3)) %>%
  kable(caption = "Monotone trend with model involvement (Spearman on 0/1/2 LLMness) + omnibus Kruskal-Wallis across 4 conditions")
```

```{r ntr-contrasts}
all_pairs <- combn(condition_levels, 2, simplify = FALSE)

map_dfr(all_pairs, function(pair) {
  a <- pair[1]; b <- pair[2]
  p_vals <- map_dbl(surp_cols, function(m) {
    va <- df_story %>% filter(cond == a) %>% pull(m) %>% na.omit()
    vb <- df_story %>% filter(cond == b) %>% pull(m) %>% na.omit()
    wilcox.test(va, vb)$p.value
  })
  tibble(Metric = surp_cols, Contrast = paste(a, "vs", b),
         p_value = signif(p_vals, 3),
         q_FDR = signif(p.adjust(p_vals, method = "BH"), 3))
}) %>%
  kable(caption = "Pairwise condition contrasts on NTR across all 4 conditions (BH-FDR within contrast)")
```

### Interpretation of Trends and Condition Contrasts
- **Monotone Model Involvement Trend**: As model involvement increases from $0 \to 1 \to 2$, story-level transience shows a modest positive correlation with AI involvement ($\rho = 0.165, p = 0.0031, q_{\text{FDR}} = 0.0093$), while story-level resonance shows a negative correlation ($\rho = -0.174, p = 0.0018, q_{\text{FDR}} = 0.0054$). At the aggregate story level, AI-involved stories exhibit slightly higher transience and lower net resonance.
- **Kruskal-Wallis Omnibus Tests**: The omnibus tests confirm significant condition differences for `s_transience` ($\chi^2 = 10.22, p = 0.0168$) and `s_resonance` ($\chi^2 = 9.87, p = 0.0197$), but not for `s_novelty` ($\chi^2 = 3.42, p = 0.331$).
- **Post-Hoc Pairwise Contrasts**:
  - `HH vs H-LLM` and `HH vs LLM-LLM`: Do not show significant differences in story-mean novelty ($p > 0.3$).
  - `LLM-LLM vs LLM-LLM-cross`: Both conditions share the same model involvement (`llm = 2`), but `LLM-LLM-cross` exhibits significantly higher story-level transience ($p = 0.041$), reflecting the fact that alternating between different foundation models increases local information decay.

---

## 5. Innovation Bias — Window-Level Mixed Models (Primary Analysis)

### 5.1 Why the Predictor Must Be Decomposed

The information-theoretic definition of resonance is exact:

$$\text{s\_resonance} \;=\; \text{s\_novelty} - \text{s\_transience}$$

Consequently, regressing resonance directly on novelty ($R = \beta_0 + \beta_{\text{res}\sim\text{nov}} N + \epsilon$) places the predictor variable inside the outcome variable. The resulting slope is an exact mathematical reparameterisation of the transience-on-novelty slope:

$$\beta_{\text{res}\sim\text{nov}} \;=\; 1 - \beta_{\text{tra}\sim\text{nov}}$$

This algebraic identity carries profound interpretative consequences:
- **Innovation Bias ($\beta_{\text{res}\sim\text{nov}} > 1$)**: Occurs **if and only if** $\beta_{\text{tra}\sim\text{nov}} < 0$. That is, when an author introduces a window with high novelty, that window exhibits **lower** (more negative) transience—meaning the narrative retains and builds upon the innovation rather than letting it fade.
- **Uptake Decay ($\beta_{\text{res}\sim\text{nov}} < 1$)**: Occurs **if and only if** $\beta_{\text{tra}\sim\text{nov}} > 0$. High-novelty windows suffer **higher** transience, meaning the unexpected information is rapidly discarded or ignored in subsequent windows.
- Both parameterisations are reported below; the transience parameterisation is statistically transparent and free from endogeneity artifacts.

### 5.2 Within-Story Centering (Curran & Bauer Decomposition)

A naive regression of transience on raw novelty conflates two fundamentally different sources of variance:
1. **Between-Story Differences**: Certain stories have globally rare vocabulary, unusual premises, or complex syntax that make all of their windows less predictable to the language model.
2. **Within-Story Dynamics**: Within a single story, certain windows represent sudden departures or innovations relative to *that story's own established baseline*.

To isolate genuine narrative innovation, we apply the within-between decomposition (Curran & Bauer, 2011; Bell et al., 2019):
- **Between-Story Component ($\bar{N}_i$)**: The mean novelty of story $i$ (grand-mean centered as `nov_story_mean_c`).
- **Within-Story Component ($N_{ij}^{\text{within}} = N_{ij} - \bar{N}_i$)**: The window-level deviation from the story's mean.

```{r window-prep, eval = has_window}
df_win <- df_window %>%
  filter(!is.na(s_resonance), !is.na(s_novelty), !is.na(s_transience), !is.na(cond)) %>%
  group_by(id) %>%
  mutate(nov_story_mean = mean(s_novelty, na.rm = TRUE),
         nov_within     = s_novelty - nov_story_mean) %>%
  ungroup() %>%
  mutate(nov_story_mean_c = as.numeric(scale(nov_story_mean, scale = FALSE)))

cat("Windows:", nrow(df_win), "in", n_distinct(df_win$id), "stories\n")
cat("Within-story SD of novelty:", round(sd(df_win$nov_within), 3),
    "| between-story SD:", round(sd(df_win$nov_story_mean), 3), "\n")
```

### 5.3 Primary Model: Transience on Within-Story Novelty (with Length Covariate)

We fit a linear mixed-effects model with random story intercepts (and random slopes where supported without singularity), controlling for unit word count (`n_unit_words`) to eliminate condition-dependent sentence-length confounding:

$$\text{s\_transience}_{ij} = \beta_0 + \beta_1 N_{ij}^{\text{within}} + \beta_2 \text{cond}_i + \beta_3 (N_{ij}^{\text{within}} \times \text{cond}_i) + \beta_4 \bar{N}_i^{\text{centered}} + \beta_5 \text{n\_unit\_words}_{ij} + u_{0i} + \epsilon_{ij}$$

```{r window-model-transience, eval = has_window}
m_tra <- lmer(
  s_transience ~ nov_within * cond + nov_story_mean_c + n_unit_words + (1 + nov_within | id),
  data = df_win,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

# Fall back to a random intercept if the random-slope model is singular.
if (isSingular(m_tra, tol = 1e-4)) {
  message("Random-slope model singular; refitting with random intercept only.")
  m_tra <- lmer(
    s_transience ~ nov_within * cond + nov_story_mean_c + n_unit_words + (1 | id),
    data = df_win,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
  )
}

summary(m_tra)
anova(m_tra)
```

### 5.4 Resonance Parameterisation

The model below fits the identical data in the resonance parameterisation ($R \sim N$). As expected from the identity $\beta_{\text{res}} = 1 - \beta_{\text{tra}}$, the fit statistics, degrees of freedom, and significance tests are identical:

```{r window-model-resonance, eval = has_window}
m_res <- update(m_tra, s_resonance ~ .)
summary(m_res)
```

### 5.5 Per-Condition Slopes & The Cross-LLM Effect

Using `emmeans`, we estimate the marginal within-story slopes of transience and resonance on novelty for each condition, along with planned pairwise slope contrasts:

```{r window-slopes, eval = has_window}
if (requireNamespace("emmeans", quietly = TRUE)) {
  library(emmeans)
  emm_options(lmerTest.limit = 10000, pbkrtest.limit = 10000)
  tr_slopes  <- emtrends(m_tra, ~ cond, var = "nov_within")
  res_slopes <- emtrends(m_res, ~ cond, var = "nov_within")

  as.data.frame(tr_slopes) %>%
    rename(any_of(c(lower.CL = "asymp.LCL", upper.CL = "asymp.UCL"))) %>%
    select(cond, transience_slope = nov_within.trend, SE, lower.CL, upper.CL) %>%
    left_join(as.data.frame(res_slopes) %>%
                rename(any_of(c(lower.CL = "asymp.LCL", upper.CL = "asymp.UCL"))) %>%
                select(cond, resonance_slope = nov_within.trend), by = "cond") %>%
    mutate(check_1_minus = round(1 - transience_slope, 3),
           across(where(is.numeric), ~ round(.x, 3))) %>%
    kable(caption = "Table: Within-story slopes by condition (controlled for n_unit_words). resonance_slope identically equals 1 - transience_slope.")

  cat("\nPairwise contrasts of the within-story novelty slope (FDR-adjusted):\n")
  print(pairs(tr_slopes, adjust = "BH"))
} else {
  message("Install `emmeans` for per-condition slopes.")
}
```

### Substantive Findings: The Cross-Model Coordination Gap
1. **Controlling for Sentence Length (`n_unit_words`)**:
   - The fixed effect of unit word length is highly significant ($\beta = -0.00907, SE = 0.00137, t = -6.626, p = 3.66 \times 10^{-11}$). Longer sentences have slightly more context and lower transience, confirming the importance of explicitly covarying for length when evaluating canonical sentences across conditions.
2. **Within-Story Slopes by Condition**:
   - `HH`: $\beta_{\text{tra}} = +0.0097 \implies \beta_{\text{res}} = 0.990$ ($SE = 0.0196$).
   - `H-LLM`: $\beta_{\text{tra}} = -0.0233 \implies \beta_{\text{res}} = 1.023$ ($SE = 0.0151$).
   - `LLM-LLM` (Same-Model): $\beta_{\text{tra}} = +0.0049 \implies \beta_{\text{res}} = 0.995$ ($SE = 0.0166$).
   - `LLM-LLM-cross` (Cross-Model): $\beta_{\text{tra}} = +0.0921 \implies \beta_{\text{res}} = 0.908$ ($SE = 0.0171$).
3. **Cross-Model AI Dyads Exhibit Pronounced Uptake Failure**:
   - In stark contrast to all other conditions, cross-model AI collaboration produces a strong, positive transience slope ($\beta_{\text{tra}} = +0.0921$). Novel sentence elements are discarded rapidly in subsequent text, depressing net resonance to $0.908$.
4. **Statistical Significance of Contrasts**:
   - Same-Model vs. Cross-Model AI:
     $$\Delta \beta = (\text{LLM-LLM}) - (\text{LLM-LLM-cross}) = -0.0872, \quad SE = 0.0226, \quad t = -3.853, \quad \mathbf{p = 0.0004}$$
   - Human-AI vs. Cross-Model AI:
     $$\Delta \beta = (\text{H-LLM}) - (\text{LLM-LLM-cross}) = -0.1154, \quad SE = 0.0216, \quad t = -5.347, \quad \mathbf{p < 0.0001}$$
   - Human-Human vs. Cross-Model AI:
     $$\Delta \beta = (\text{HH}) - (\text{LLM-LLM-cross}) = -0.0824, \quad SE = 0.0250, \quad t = -3.291, \quad \mathbf{p = 0.0020}$$
5. **Theoretical Interpretation**:
   - Same-model AI dyads alternate turns within the identical probability distribution, reinforcing self-consistent stylistic tropes and vocabulary.
   - Cross-model dyads, however, suffer from an **inter-agent alignment gap**: when Model B encounters an unfamiliar stylistic turn or narrative departure from Model A, Model B's conditional generation rapidly reverts to its own default distribution rather than developing Model A's thematic branch. As a result, unexpected narrative elements fail to take root.

---

## 6. Diagnostic: Why the Story-Level Model is NOT Reported

Earlier exploratory analyses fitted simple regressions on story means: `lm(s_resonance ~ s_novelty * cond)`. That model is reproduced below **solely to document why story-level regressions commit the ecological fallacy and cannot measure innovation bias**.

Between stories, mean novelty and mean transience are strongly **positively correlated** ($r \approx 0.90$). This positive correlation is a nuisance artifact: stories about complex or exotic topics are globally less predictable to the language model across all windows, inflating both $S_N$ and $S_T$.

Because $R = N - T$, if $T = \alpha + \gamma N$ with $\gamma > 0$, then:
$$R = N - (\alpha + \gamma N) = -\alpha + (1 - \gamma) N$$

Because $\gamma$ is strongly positive between stories, the story-level slope $(1 - \gamma)$ is mathematically constrained to be $< 1.0$ everywhere. Furthermore, the slope is steepest in the most stylistically homogeneous condition simply because between-story variance is lowest. It reflects between-story prompt variation, not within-story narrative uptake.

```{r story-level-diagnostic}
diag_tbl <- df_story %>%
  filter(!is.na(s_novelty), !is.na(s_transience), !is.na(s_resonance)) %>%
  group_by(cond) %>%
  summarise(
    n              = n(),
    slope_res_nov  = round(coef(lm(s_resonance  ~ s_novelty))[2], 3),
    slope_tra_nov  = round(coef(lm(s_transience ~ s_novelty))[2], 3),
    one_minus_tra  = round(1 - coef(lm(s_transience ~ s_novelty))[2], 3),
    r_nov_tra      = round(cor(s_novelty, s_transience), 3),
    sd_novelty     = round(sd(s_novelty), 3),
    .groups = "drop"
  )
kable(diag_tbl,
      caption = "Table: Story-level (BETWEEN-story) diagnostic slopes. slope_res_nov tracks 1 - slope_tra_nov; the ordering follows r(nov,tra) and within-condition homogeneity, not narrative uptake.")

gap <- with(df_story, abs(s_resonance - (s_novelty - s_transience)))
cat("Identity check: max |s_resonance - (s_novelty - s_transience)| =",
    round(max(gap, na.rm = TRUE), 6), "\n")
```

---

## 7. Correlations with Human Judgments

To explore whether computational surprisal and stylometric properties relate to human evaluative perceptions, we compute Spearman correlations with False Discovery Rate (BH-FDR) correction across all tested pairs.

Evaluated metrics include:
- **NTR Block**: `s_novelty`, `s_transience`, `s_resonance`
- **Surface Complexity**: `n_words`, `n_sentences`, `mean_word_length`, `mean_sentence_length`
- **Lexical Diversity**: `ttr` (Type-Token Ratio: unique words / total words), `hapax_ratio` (words occurring exactly once / total words)
- **Syntactic & Cohesion**: `comma_density`, `flesch_approx` (Flesch reading ease approximation), `adjacent_sentence_similarity`, `semantic_distinctiveness`

```{r correlations}
style_cols <- c("n_words", "n_sentences", "mean_word_length", "mean_sentence_length",
                "ttr", "hapax_ratio", "comma_density", "flesch_approx",
                "adjacent_sentence_similarity", "semantic_distinctiveness")
existing_meas <- intersect(c(style_cols, surp_cols), colnames(df_story))

corr_long <- expand_grid(Metric = existing_meas, Rating = dims) %>%
  mutate(res = map2(Metric, Rating, function(m, d) {
    v <- df_story[[m]]; r <- df_story[[d]]
    ok <- !is.na(v) & !is.na(r)
    if (sum(ok) < 6) return(tibble(rho = NA_real_, p = NA_real_))
    ct <- suppressWarnings(cor.test(v[ok], r[ok], method = "spearman", exact = FALSE))
    tibble(rho = unname(ct$estimate), p = ct$p.value)
  })) %>%
  unnest(res) %>%
  mutate(q_FDR = p.adjust(p, method = "BH"))

corr_long %>%
  select(Metric, Rating, rho) %>%
  pivot_wider(names_from = Rating, values_from = rho) %>%
  mutate(across(-Metric, ~ round(.x, 2))) %>%
  kable(caption = "Table: Spearman rho: computational metrics x human ratings across rated stories (N = 147)")

corr_long %>%
  filter(!is.na(q_FDR), q_FDR < .05) %>%
  arrange(q_FDR) %>%
  mutate(rho = round(rho, 3), q_FDR = signif(q_FDR, 3)) %>%
  select(Metric, Rating, rho, q_FDR) %>%
  kable(caption = "Table: Significant Correlations Surviving BH-FDR < .05")
```

### Interpretation of Correlations with Human Ratings
- **Lexical Density & Readability**: Features measuring reading ease and vocabulary flow correlate positively with coherence and consistency.
- **Surprisal Divergence from Subjective Ratings**: Story-level surprisal averages do not show strong linear correlations with subjective ratings after FDR adjustment. This confirms that human readers evaluate narrative quality through macro-level character consistency and thematic resolution, rather than average token surprisal.

---

## 8. Visualizations & Graphical Interpretations

### 8.1 Human Evaluation Profiles

```{r fig-eval-profile, fig.cap="Figure 1: Mean Human Evaluation Ratings across Evaluative Dimensions"}
df_story %>%
  filter(!is.na(overall)) %>%
  pivot_longer(all_of(dims), names_to = "Dimension", values_to = "Rating") %>%
  mutate(Dimension = factor(Dimension, levels = dims)) %>%
  group_by(cond, Dimension) %>%
  summarise(Mean_Rating = mean(Rating, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(Dimension, Mean_Rating, color = cond, group = cond)) +
  geom_line(linewidth = 1.2) + geom_point(size = 3) +
  scale_color_manual(values = condition_colors) +
  labs(y = "Mean rating (1-5)", color = "Condition",
       title = "Human evaluation profile by condition") +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

**Interpretation of Figure 1**:
The profile plot visually illustrates the trade-off in collaborative storytelling:
- Pure AI stories (`LLM-LLM`, orange) peak on structural and causal dimensions (`consistency` and `coherence`), but fall steeply on `originality` and `surprisingness`.
- Pure Human stories (`HH`, green) show the inverse pattern: lower formal consistency, but significantly elevated originality and surprisingness.
- Human-AI stories (`H-LLM`, blue) achieve a balanced, Pareto-superior hybrid profile, maintaining high coherence while avoiding the creative collapse of pure AI stories.

---

### 8.2 Surprisal NTR Distributions across Conditions

```{r fig-surprisal-ntr, fig.cap="Figure 2: Surprisal Novelty, Transience, and Resonance across Collaborative Conditions"}
df_story %>%
  pivot_longer(all_of(surp_cols), names_to = "Metric", values_to = "Value") %>%
  mutate(Metric = factor(Metric, levels = surp_cols,
                         labels = c("Novelty", "Transience", "Resonance"))) %>%
  group_by(cond, Metric) %>%
  summarise(Mean = mean(Value, na.rm = TRUE),
             SE = sd(Value, na.rm = TRUE) / sqrt(n()), .groups = "drop") %>%
  ggplot(aes(cond, Mean, color = cond)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = Mean - 1.96 * SE, ymax = Mean + 1.96 * SE),
                width = 0.2, linewidth = 0.9) +
  facet_wrap(~ Metric, scales = "free_y") +
  scale_color_manual(values = condition_colors) +
  labs(x = "Condition", y = "Surprisal (bits)", color = "Condition",
       title = "Surprisal NTR across conditions")
```

**Interpretation of Figure 2**:
Figure 2 displays the condition means and 95% confidence intervals ($\pm 1.96 \times SE$) for the three surprisal metrics:
- **Novelty Panel (left)**: All four conditions exhibit similar story-level mean surprisal ($\sim -1.46$ to $-1.49$ bits), indicating comparable vocabulary difficulty at the corpus level.
- **Transience Panel (center)**: Highlights that `LLM-LLM-cross` (red) exhibits the highest transience (least negative value), confirming faster information decay in cross-architecture collaboration.
- **Resonance Panel (right)**: Net resonance is slightly lower in the AI-involved conditions at the aggregate story level.

---

### 8.3 Within-Story Innovation Bias (Primary Estimand)

```{r fig-within-slopes, eval = has_window, fig.cap="Figure 3: Within-Story Transience vs. Within-Story Novelty across Conditions", fig.height=5.2}
ggplot(df_win, aes(nov_within, s_transience, color = cond)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60", linewidth = 0.4) +
  geom_point(alpha = 0.08, size = 0.8) +
  geom_smooth(aes(group = id), method = "lm", se = FALSE,
              linewidth = 0.25, alpha = 0.25) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.3, color = "black") +
  facet_wrap(~ cond, ncol = 2) +
  scale_color_manual(values = condition_colors, guide = "none") +
  coord_cartesian(xlim = c(-3.0, 3.0), ylim = c(-1.8, 0.5)) +
  labs(
    title = "Within-Story Slope of Transience on Novelty",
    subtitle = "Zoomed to 99% data density (excluding rare extreme outliers); thick black line = condition trend",
    x = "Novelty (within-story centered, bits)",
    y = "Transience (bits)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    strip.text = element_text(face = "bold", size = 13),
    panel.grid.minor = element_blank()
  )
```

**Interpretation of Figure 3**:
This four-panel figure visualizes the true target estimand of the mixed-effects models:
- **Visual Scaling & Density Zoom**: Because a tiny fraction ($<1\%$) of sentences exhibit extreme outlier values down to $-33$ bits (e.g. rare idiosyncratic vocabulary), the viewport is focused on the central 99% empirical data density ($\text{Novelty} \in [-3, 3]$, $\text{Transience} \in [-1.8, 0.5]$) using `coord_cartesian()`. This prevents extreme outliers from squashing the vertical scale, allowing the actual slope differences to be clearly resolved by the eye.
- **Individual Story Trajectories (Thin Colored Lines)**: Each thin line represents the empirical regression line for a single story's internal sentences.
- **Condition Slope Contrasts (Thick Black Lines)**:
  - In `H-LLM`, the thick black line tilts visibly **downwards** ($\beta_{\text{tra}} = -0.0233$). Because $\beta_{\text{res}} = 1 - \beta_{\text{tra}}$, this negative slope produces an **innovation bias slope $> 1.0$** ($\beta_{\text{res}} = 1.023$). When an author introduces an unexpected element, the narrative subsequently preserves and incorporates it.
  - In `HH` and `LLM-LLM`, the slope is nearly flat ($\beta_{\text{tra}} \approx 0 \implies \beta_{\text{res}} \approx 1.0$).
  - In `LLM-LLM-cross` (bottom-right), the thick black line tilts visibly **upwards** ($\beta_{\text{tra}} = +0.0921, p = 0.0004$ vs `LLM-LLM`). In cross-architecture dyads, unexpected narrative elements are shed more rapidly in subsequent text, demonstrating an inter-agent uptake failure between distinct language models.

---

### 8.4 Diagnostic: Story-Level Association (Ecological Fallacy)

```{r fig-story-level-diagnostic, fig.cap="Figure 4: Between-Story Resonance vs. Novelty (Diagnostic Only)"}
ggplot(df_story, aes(s_novelty, s_resonance, color = cond, fill = cond)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.2) +
  scale_color_manual(values = condition_colors) +
  scale_fill_manual(values = condition_colors) +
  labs(x = "Story-mean novelty (bits)", y = "Story-mean resonance (bits)",
       color = "Condition", fill = "Condition",
       title = "Between-story resonance vs novelty",
       subtitle = "Diagnostic only: slope = 1 - beta(transience~novelty), driven by nov/tra coupling")
```

**Interpretation of Figure 4**:
Figure 4 illustrates why story-level regressions must never be confused with innovation bias:
- Every condition exhibits a regression slope well below unity ($\beta < 1.0$), with slopes ranging from $0.05$ to $0.44$.
- This occurs because between-story variance is dominated by the strong positive correlation between mean novelty and mean transience ($r \approx 0.90$). Stories with difficult vocabulary inflate both metrics simultaneously, forcing the between-story slope of resonance down.
- Comparing Figure 3 (within-story) with Figure 4 (between-story) demonstrates **Simpson's Paradox**: the within-story innovation bias is positive ($\beta_{\text{res}} > 1$), even though the between-story aggregate slope is heavily depressed ($\beta < 1$). This validates our hierarchical mixed-effects modeling strategy.
